A Geometry for Network Visualization in \MakeLowercase{\texttt{ggplot2}}

### 2015-10-05 inst/paper.Rnw
including geom_net version of examples into the paper

\maketitle
\begin{abstract}
There are many implementations of static network visualization in R, but none are equipped with the flexibility and functionality of \texttt{ggplot2}.  \texttt{geom\_net} was created to fill this gap. Using two data frames to describe vertex and edge information, \texttt{geom\_net} makes use of the underlying structure of \texttt{ggplot2} to visualize networks. This makes it possible to easily facet networks according to covariates  or change aesthetics such as shape, size, or color according to additional edge or vertex information.
\end{abstract}
<<chunk_setup, echo=FALSE>>=
library(knitr)
opts_chunk$set(tidy = FALSE, echo=FALSE, cache=FALSE, eval = TRUE, message = FALSE, warning = FALSE,  fig.height = 8, fig.width = 8, fig.align = 'center',fig.show='hide')
@
<<readingdata, eval=TRUE, echo=FALSE>>=
library(ggplot2) # needs to be the version with geom_net!
 theme_net <- theme(aspect.ratio = 1, line = element_blank(), panel.background = element_blank(), plot.background = element_blank(), axis.text=element_blank(),axis.title=element_blank())
 data(blood, football, theme_elements, lesmis,madmen, mm_directed,email)
names(blood$vertices)[1] <- 'label'
#load("data/madmen.rda") # relative path should work for both of us
#load("data/mm_directed.rda")
#names(madmen$vertices)[1] <- 'label'
names(theme_elements$vertices)[1] <- 'label'
football$vertices$label <- as.factor(football$vertices$label)
football$vertices$value <- as.factor(football$vertices$value)
football$edges$in.conf <- 0
for (i in 1:nrow(football$edges)){
  idx1 <- which(football$vertices$label == football$edges$from[i])
  idx2 <- which(football$vertices$label == football$edges$to[i])
  if (football$vertices$value[idx1] == football$vertices$value[idx2]){
    football$edges$in.conf[i] <- 1
  }
}
#load("data/email.rda")
#email.net <- read.csv("data/email-data.csv")
#email.names <- data.frame(label = as.factor(unique(c(as.character(email.net$From),as.character(email.net$to)))))
#library(plyr)
#email.net <- ddply(email.net, .(eID), transform, nrecipients=length(to))
#employee <- read.csv("data/EmployeeRecords.csv")
#employee <- merge(email.names, employee, by.x="label", by.y="EmailAddress", all.x=T)
email.net <- email$edges
employee <- email$nodes
library(RColorBrewer)
cols <- brewer.pal(8, name="Set1")
load("data/protein.rda")

#yeast <- list(edges = protein, vertices = data.frame(label = min(yeast):max(yeast)))
yeast <- protein
@

\section{Introduction}
At its core, a network is simply a set of points connected in pairs by a set of lines \citep{newman}.  Here, we refer to the lines as edges and the points as vertices, although these are also called nodes.  These two seemingly simple sets of graphical objects, points and segments, are used to encode a huge variety and quantity of information across many fields of study. For instance, networks of scientific collaboration, a food web of marine animals, and American college football games are all covered in a paper on community detection in networks by \citet{football}.  Additionally, \citet{networkfailures} examine node failure in interdependent networks like power grids.  Social networks, such as links between actors found on \url{www.imdb.com}, and neural networks, like the completely mapped neural network of the \textit{C. elegans} worm are also etensively studied \citep{smallworld}.  Networks vary widely in scope and complexity: the smallest network is simply an edge between two vertices, while one of the most commonly used and most complex networks, the world wide web, has billions of vertices (webpages) and billions of edges (hyperlinks) connecting them.  The edges in a network can  be directed or undirected: directed edges represent information travelling from one vertex to another, and switching the direction would change the structure of the network. The world wide web is an example of a directed network because one webpage may link to another, and not necessarily the other way around. Undirected edges, however, are simply connections between vertices.  Coauthorship networks that encode information about academic publications are examples of undirected networks because if two people author a paper together, that creates a connection between them that is bidirectional.

@@

This network is shown in figure~\ref{fig.cap:madmen_ex}.
<<madmen_ex, dependson='madmen-prep', fig.cap = 'Network of Relationships in Mad Men'>>=
ggplot(data = madmen$edges, aes(from_id = Name1, to_id = Name2)) +
  geom_net(vertices = madmen$vertices, vsize=3, vlabel= TRUE, ecolour="grey30",
           aes(vcolour=c( "#FF69B4", "#0099ff")[as.numeric(Gender)])) +
  expand_limits(x = c(0,1.25), y = c(0,1)) + theme_net
@

\begin{figure}
\centering
\includegraphics[width=\textwidth]{figure/madmen_ex-1.pdf}
\caption{\label{fig.cap:madmen_ex} Graph of the characters in the show Mad Men who are linked by a romantic relationship.}
\end{figure}
\afterpage{\clearpage}

@@

  In our geometry, we want to create all possible vertex aesthetic to mimic the usual aesthetics in \texttt{ggplot2} for points. The \texttt{vcolor} aesthetic can be an identity color (e.g. \texttt{color = "red"}) to change the color of all vertices, or, in the final version, it will take a factor variable which identifies each vertex as belonging to one of several groups, and it colors the vertices of the graph according to this grouping. In the current version, the \texttt{vcolor} aesthetic does not work like a typical color aesthetic in \texttt{ggplot2}. It can, however, take HEX color characters and assign them to different levels of a factor variable. We use the \texttt{RColorBrewer} package to choose the colors for us in this paper \citep{RColorBrewer}. It will also color the vertices along a color gradient according to the different values of some numeric variable such as degree.  The vertex color defaults to black. % are set to the default color palette in \texttt{ggplot2}, and the palette used can be changed in the usual way within the \texttt{ggplot2} framework, by using \texttt{scale\_color\_brewer}.}
  %The \texttt{vcolor} aesthetic operates in the same way as the \texttt{vfill} aesthetic, but it changes the outline color of the vertex instead of the fill color of the vertex, which is in line with the \texttt{ggplot2} use of the \texttt{color} and \texttt{fill} aesthetics.
  The \texttt{vsize} aesthetic can also be set to an identity value to increase or decrease the size of all vertices in the graph, or it can be set to one of any numerical columns in the vertex data frame. Right now, the numerical values are strict point values: if the size of a vertex is two, then it will be two points in diameter. A point is the size of the default fontsize in the \texttt{grid} package. In our final version, the \texttt{vsize} will scale to minimum and maximum values as all other size aesthestics in \texttt{ggplot2} do.  The \texttt{vshape} aesthetic is for changing the shape of the vertices from the default circle to other shapes, like square or triangle.  It also changes the shape of the vertices according to some grouping variable, which is currently required to consist of integer values from 0 to 25.  In the final version, it will be able to change shape based on different levels of a factor variable.  Finally, the \texttt{valpha} aesthetic will change the vertex transparency to a set value, like 1/10 or 0.5.  It operates in exactly the same way as the \texttt{alpha} aesthetic in \texttt{geom\_point}.  This aesthetic is very useful for networks with hundreds or thousands of vertices, which can easily crowd the static visualization. Making the vertices more transparent will better show the underlying structure in the network.

@@

\section{Usage Examples}
In this section, we demonstrate the current capabilities of \texttt{geom\_net} in a series of diverse examples.
  \subsection{Blood Donation}
  In this directed network, there are eight vertices and 27 edges.  The vertices represent the eight different blood types in humans that are most important for donation: the ABO blood types A, B, AB, and O, combined with the RhD positive (+) and negative (-) types. The edges are directed: a person whose blood type is that of a \emph{from} vertex can to donate blood to a person whose blood type is that of a corresponding \emph{to} vertex. In the example below, loops are removed because loops exist on every vertex in this example, as blood between two people of matching ABO and RhD type can always be exchanged.
<<blood_ex, echo=TRUE, fig.cap= "Directed network of blood type donations in humans.">>=
ggplot(data = blood$edges, aes(from_id = from, to_id = to)) +
  geom_net(vertices = blood$vertices, vcolour = I('red'), layout = 'circle',
           vlabel = TRUE, vsize = I(3), directed = TRUE) +
  expand_limits(x = c(0,1), y = c(0,1)) + theme_net
@
This network is shown in figure \ref{fig.cap:blood_ex}.  Here, we have used the aesthetics \texttt{vcolour} and \texttt{vsize} set to identity values to change the size and color of all vertices. We have also used the \texttt{layout} and \texttt{vlabel} arguments to change the default layout to a circle layout and to print the blood types, respectively. The circle layout places blood types of the same ABO type next to each other and spreads the vertices out far enough to distinguish between the various ``in" and ``out" types.  You can tell clearly from this plot that the O- type is the universal donor: it has out degree of seven and in degree of zero. Additionally, you can see that the AB+ type is the universal recepient, with in degree of seven and out degree of zero. Anyone looking at this plot can quickly determine which type(s) of blood they can receive and which type(s) can receive their blood.
\begin{figure}
\centering
\includegraphics[width=\textwidth]{figure/blood_ex-1.pdf}
\caption{\label{fig.cap:blood_ex} Network of blood donation possibilities in humans by ABO and RhD blood types.}
\end{figure}
\afterpage{\clearpage}

\subsection{Email Network}
  This email network comes from the 2014 VAST Challenge \citep{emailnet}. It is a directed network of emails between company employees with 55 vertices and 9,063 edges. Each vertex is an employee of the company, and each edge is an email sent from one employee to one or more other employees. The arrow of the directed edge points to the recipient(s) of the email. The network contains two business weeks of emails across the entire company. In order to better visualize the structure of the communication network between employees, all emails that were sent out to all employees are removed in the subsequent examples.
<<email_ex, fig.cap = 'All emails within a company over a two week period.',echo=TRUE>>=
ggplot(data = subset(email.net, nrecipients < 54), aes(from_id = From, to_id = to)) +
  geom_net(vertices = employee,ealpha = 0.1, vsize = I(4), directed = TRUE,
           aes(vcolour=cols[as.numeric(CurrentEmploymentType)])) +
  expand_limits(x=c(0,1), y = c(0,1)) + theme_net
@
\begin{figure}
\centering
\includegraphics[width=\textwidth]{figure/email_ex-1.pdf}
\caption{\label{fig.cap:email_ex} Email network within a company over a two week period.}
\end{figure}
\afterpage{\clearpage}

@@

<<email_ex_facet, fig.height=8, fig.width=8, fig.cap = 'All emails within a company over a two week period facetted by day of the month.',echo=TRUE,eval=FALSE>>=
ggplot(data = subset(email.net, nrecipients < 54), aes(from_id = From, to_id = to)) +
  geom_net(vertices = employee, ealpha = 0.2, vsize = I(2), directed = TRUE,
           aes(vcolour=cols[as.numeric(CurrentEmploymentType)])) +
  expand_limits(x=c(0,1), y = c(0,1)) + facet_wrap(~day, nrow = 2) + theme_net

@@

\includegraphics[width=8in]{figure/email_ex_facet-1.pdf}

@@

This plot is shown in figure \ref{fig.cap: email_ex_facet}. With the facetting, we can see that there are several days where one or more departments do not communicate with any of the other departments. There are only two days, 13 and 15, without any isolated department communications. Facetting is one of the major benefits of creating a geometry for networks in \texttt{ggplot2}.  Facetting quickly separates dense networks into separate subnetworks for easy visual comparison and  analyses.
  \subsection{\texttt{ggplot2} Theme Elements}
This example comes from the \texttt{theme()} help page in the \texttt{ggplot2} documentation \citep{ggplot}.  It is a directed network which shows the structure of the inheritance of theme options in the construction of a \texttt{ggplot2} plot.  There are 53 vertices and 36 edges in this network. Each vertex represents one possible theme option. There is an arrow from one theme option to another if the element represented by the \emph{to} vertex inherits its values from the \emph{from} vertex.  For example, the \texttt{axis.ticks.x} option inherits its value from the \texttt{axis.ticks} value, which inherits its value from the \texttt{line} option.  Thus, setting the \texttt{line} option to a value such as \texttt{element\_blank()} sets the entire inheritance tree to \texttt{element\_blank()}, and no lines appear anywhere on the plot background. Finally, we note that the vertices with no edges were incorporated into the plot by adding their labels to the edges data frame in both the  `from\_id' and `to\_id' columns before passing the edges data frame to \texttt{ggplot}.
<<theme_ex, echo=TRUE, fig.cap= 'Inheritance structure for \\texttt{theme} elements in \\texttt{ggplot2}.'>>=
ggplot(data = theme_elements$edges, aes(from_id = parent, to_id = child)) +
  geom_net(vertices = theme_elements$vertices, directed = TRUE, vlabel = TRUE) +
  expand_limits(x = c(0,1.2), y = c(0,1)) +
  theme_net
\includegraphics[width=\textwidth]{figure/theme_ex-1.pdf}
\caption{\label{fig.cap:theme_ex} Inheritance structure of \texttt{ggplot2} theme elements. This is a recreation of the graph found at \protect\url{http://docs.ggplot2.org/current/theme.html}.}
\afterpage{\clearpage}

The inheritance structure is plotted in figure \ref{fig.cap:theme_ex}.  In this plot, it is easy to quickly determine the parent and child vertices. Using this plot made creation of the \texttt{theme\_net} object used throughout these examples very simple. We just made set each of the major parent elements, \texttt{text}, \texttt{rect}, and \texttt{line} to \texttt{element\_blank()} and then set the aspect ratio equal to one.
  \subsection{Mad Men Networks}
  The following code creates the network example given in the introduction. We changed the vertex size and edge color for all vertices and edges, included vertex labels, and colored the vertices according to the character's gender.
<<print_ex, ref.label="madmen_ex", eval=FALSE, echo=TRUE,cache=FALSE>>=
@
 There is another Mad Men network included in the \texttt{gcookbook} package \citep{madmen}. It is a directed network, also of romantic relationships between characters, but it also includes advances made by one character that were rejected by the other.  For example, Roger Sterling made advances toward Betty Draper, but Betty refused him, and so there is a directed edge going from Roger to Betty, but not from Betty to Roger.  If the advance is reciprocated, like between Sal Romano and the Bellhop, there are two directed edges between the two vertices.
<<madmen2_ex,echo=TRUE, fig.cap= 'A directed network of relationships in Mad Men.'>>=
ggplot(data = mm.directed$edges, aes(from_id = Name1, to_id = Name2)) +
  geom_net(vertices = mm.directed$vertices, directed = T, vlabel = T,
           vsize = I(2.5), layout = 'fruchtermanreingold',
           aes(vcolour=c( "#FF69B4", "#0099ff")[as.numeric(Gender)])) +
  expand_limits(x = c(0,1.1), y = c(0,1)) + theme_net
@
\begin{figure}
\centering
\includegraphics[width=\textwidth]{figure/madmen2_ex-1.pdf}
\caption{\label{fig.cap:madmen2_ex} A directed network of relationships in Mad Men.}
\end{figure}
This network is shown in figure \ref{fig.cap:madmen2_ex}.  This network is much more connected than the previous Mad Men example.  It also allows us to see much more of the drama in the show. For instance, you can see that Roger made advances towards Betty, his business partner's wife, which is much more scandalous than what we can see in the first network.

  \subsection{College Football}
  This next example comes from M.E.J. Newman's network data web page \citep{football}.  It is an undirected network consisting of all regular season college football games played between Division I schools in Fall of 2000.  There are 115 vertices and 613 edges: each vertex represents a school, and an edge represents a game played between two schools. There is an additional variable in the vertex data frame corresponding to the conference each team belongs to, and there is an additional variable in the edge data frame that is equal to one if the game occured between teams in the same conference or zero if the game occured between teams in different conferences.
<<football_ex, echo=TRUE, fig.cap="All regular season Division I college football games in the Fall 2000 season.">>=
colors <- brewer.pal(12, name = 'Paired')
ggplot(data = football$edges, aes(from_id = from, to_id = to,
                                  elinetype = in.conf + 1)) +
  geom_net(vertices = football$vertices, ealpha = 0.3, vsize = 2, vlabel = TRUE,
           aes(vcolour=colors[as.numeric(factor(value))])) +
  expand_limits(x = c(0,1), y = c(0,1)) + theme_net
@
The network of football games is given in figure \ref{fig.cap:football_ex}. Here, we have changed the \texttt{elinetype} aesthetic to correspond to games that occur between teams in the same conference or different conferences.  These lines are dotted and solid, respectively. We have also assigned a different color to each conference, and the vertices and their labels are colored according to their conference. This coloring and changing of the linetypes make the structure of the game network easier to view.  There is one conference consisting of Navy, Notre Dame, Utah State, Central Florida, and Connecticut, which is spread out all over the network, whereas most other conferences' teams are all very close to each other because they play within conference much more than they play out of conference.  At the time, these five schools were all independents and did not have a home conference.  Without the coloring capability, we would not have been able to pick out that difference as easily.
\begin{figure}
\centering
\includegraphics[width=\textwidth]{figure/football_ex-1.pdf}
\caption{\label{fig.cap:football_ex} The network of regular season Division I college football games in the season of fall 2000. The vertices and their labels are colored by conference.}
\end{figure}
\afterpage{\clearpage}

  \subsection{Les Mis\'{e}rables}
  This next network comes from \citet{lesmis}. It is an undirected network of coappearances of characters in Victor Hugo's \emph{Les Mis\'{e}rables}.  There are 77 vertices representing each of the 77 characters in the book, and an edge connects two vertices if those two characters appear in the same chapter of the book.  There are 254 edges in this network. The edges are also weighted by the number of coappearances. The largest weighting is 31, between the characters Jean Valjean and Cosette.  This network is shown in figure \ref{fig.cap:lesmis_ex}. This network reveals many important character groups, as well as the main characters in the story.  The main characters are the ones with high degree, like Valjean, Cosette, and Javert.  There is an extremely wide edge connecting Valjean and Cosette, which implies that they are strongly connected; in fact, Cosette is Valjean's adopted daughter.
<<lesmis_ex, echo=TRUE, fig.cap="Coappearance network of characters in Victor Hugo's \\emph{Les Mis\\'{e}rables}">>=
ggplot(data = lesmis$edges, aes(from_id = from, to_id = to,
                                esize = value/mean(value))) +
  geom_net(vertices = lesmis$vertices, ealpha = 7/10,
           vcolour = I("red"), vsize = I(2.5),
           layout = 'fruchtermanreingold', vlabel=TRUE) +
  expand_limits(x = c(0,1.1), y = c(0,1)) + theme_net
@
\begin{figure}
\centering
\includegraphics[width=\textwidth]{figure/lesmis_ex-1.pdf}
\caption{\label{fig.cap:lesmis_ex} Co-appearance network of characters in Victor Hugo's \emph{Les Mis\'{e}rables}.}
\end{figure}
\afterpage{\clearpage}

>>>>

\maketitle
\begin{abstract}
There are many implementations of static network visualization in R, but none are equipped with the flexibility and functionality of \texttt{ggplot2}.  \texttt{geom\_net} was created to fill this gap. Using two data frames to describe vertex and edge information, \texttt{geom\_net} makes use of the underlying structure of \texttt{ggplot2} to visualize networks. This makes it possible to easily facet networks according to covariates  or change aesthetics such as shape, size, or color according to additional edge or vertex information.
\end{abstract}
<<chunk_setup, echo=FALSE>>=
library(knitr)
opts_chunk$set(tidy = FALSE, echo=FALSE, cache=FALSE, eval = TRUE, message = FALSE, warning = FALSE,  fig.height = 8, fig.width = 8, fig.align = 'center',fig.show='hide')
@
<<readingdata, eval=TRUE, echo=FALSE>>=
library(ggplot2) # needs to be version ≥ 1.0.1.9003
library(ggnet) # install from github
data(blood, football, theme_elements, lesmis,madmen, mm_directed, email)

# data(protein) # uncomment it once protein.rda is in data folder
library(RColorBrewer)
cols <- brewer.pal(8, name="Set1")
yeast <- protein

# set up networks for all of the data sets
MMnet <- merge(madmen$edges, madmen$vertices, by.x="Name1", by.y="label", all=TRUE)
bloodnet <- merge(blood$edges, blood$vertices, by.x="from", by.y="label", all=TRUE)

employee <- data.frame(expand.grid(
              label=unique(email$nodes$label), day=unique(email$edges$day)))
employee <- merge(employee, email$nodes, by="label")
emailnet <- merge(subset(email$edges, nrecipients < 54), employee, by.x=c("From", "day"), by.y=c("label", "day"), all=TRUE)

\hh{todo list:
\begin{enumerate}
\item there is no protein.rda in the data folder - could you please put it there?
\item include installation instruction for package ggnet
\end{enumerate}
}
\section{Introduction}
At its core, a network is simply a set of points connected in pairs by a set of lines \citep{newman}.  Here, we refer to the lines as edges and the points as vertices, although these are also called nodes.  These two seemingly simple sets of graphical objects, points and segments, are used to encode a huge variety and quantity of information across many fields of study. For instance, networks of scientific collaboration, a food web of marine animals, and American college football games are all covered in a paper on community detection in networks by \citet{football}.  Additionally, \citet{networkfailures} examine node failure in interdependent networks like power grids.  Social networks, such as links between actors found on \url{www.imdb.com}, and neural networks, like the completely mapped neural network of the \textit{C. elegans} worm are also etensively studied \citep{smallworld}.  Networks vary widely in scope and complexity: the smallest network is simply an edge between two vertices, while one of the most commonly used and most complex networks, the world wide web, has billions of vertices (webpages) and billions of edges (hyperlinks) connecting them.  The edges in a network can  be directed or undirected: directed edges represent information travelling from one vertex to another, and switching the direction would change the structure of the network. The world wide web is an example of a directed network because one webpage may link to another, and not necessarily the other way around. Undirected edges, however, are simply connections between vertices.  Coauthorship networks that encode information about academic publications are examples of undirected networks because if two people author a paper together, that creates a connection between them that is bidirectional.

@@

This network is shown in figure~\ref{fig.cap:madmen_ex}.

\begin{figure}[hbtp]
\centering
<<madmen_ex, dependson='readingdata', fig.cap = 'Network of Relationships in Mad Men', out.width='\\textwidth', fig.width=9, fig.height=9>>=
ggplot(data = MMnet, aes(from_id = Name1, to_id = Name2)) +
  geom_net(aes(colour=Gender), size=4, label= TRUE, vjust=-0.6,
           ecolour="grey50") +
  theme_net() +
  scale_colour_manual(values=c( "#FF69B4", "#0099ff")) +
  xlim(c(-0.05, 1.05)) + theme(legend.position="bottom")
@
\includegraphics[width=\textwidth]{figure/madmen_ex-1.pdf}
\caption{\label{fig.cap:madmen_ex} Graph of the characters in the show Mad Men who are linked by a romantic relationship.}
\end{figure}
\afterpage{\clearpage}

@@

  In our geometry, we want to create all possible vertex aesthetic to mimic the usual aesthetics in \texttt{ggplot2} for points. The \texttt{vcolor} aesthetic can be an identity color (e.g. \texttt{color = "red"}) to change the color of all vertices, or, in the final version, it will take a factor variable which identifies each vertex as belonging to one of several groups, and it colors the vertices of the graph according to this grouping. In the current version, the \texttt{vcolor} aesthetic does not work like a typical color aesthetic in \texttt{ggplot2}. It can, however, take HEX color characters and assign them to different levels of a factor variable. We use the \texttt{RColorBrewer} package to choose the colors for us in this paper \citep{RColorBrewer}. It will also color the vertices along a color gradient according to the different values of some numeric variable such as degree.  The vertex color defaults to black. % are set to the default color palette in \texttt{ggplot2}, and the palette used can be changed in the usual way within the \texttt{ggplot2} framework, by using \texttt{scale\_color\_brewer}.}
 %The \texttt{vcolor} aesthetic operates in the same way as the \texttt{vfill} aesthetic, but it changes the outline color of the vertex instead of the fill color of the vertex, which is in line with the \texttt{ggplot2} use of the \texttt{color} and \texttt{fill} aesthetics.

The \texttt{vsize} aesthetic can also be set to an identity value to increase or decrease the size of all vertices in the graph, or it can be set to one of any numerical columns in the vertex data frame. Right now, the numerical values are strict point values: if the size of a vertex is two, then it will be two points in diameter. A point is the size of the default fontsize in the \texttt{grid} package. In our final version, the \texttt{vsize} will scale to minimum and maximum values as all other size aesthestics in \texttt{ggplot2} do.  The \texttt{vshape} aesthetic is for changing the shape of the vertices from the default circle to other shapes, like square or triangle.  It also changes the shape of the vertices according to some grouping variable, which is currently required to consist of integer values from 0 to 25.  In the final version, it will be able to change shape based on different levels of a factor variable.  Finally, the \texttt{valpha} aesthetic will change the vertex transparency to a set value, like 1/10 or 0.5.  It operates in exactly the same way as the \texttt{alpha} aesthetic in \texttt{geom\_point}.  This aesthetic is very useful for networks with hundreds or thousands of vertices, which can easily crowd the static visualization. Making the vertices more transparent will better show the underlying structure in the network.

@@

\section{Usage Examples}
In this section, we demonstrate the current capabilities of \texttt{geom\_net} in a series of diverse examples.
  \subsection{Blood Donation}
  In this directed network, there are eight vertices and 27 edges.  The vertices represent the eight different blood types in humans that are most important for donation: the ABO blood types A, B, AB, and O, combined with the RhD positive (+) and negative (-) types. The edges are directed: a person whose blood type is that of a \emph{from} vertex can to donate blood to a person whose blood type is that of a corresponding \emph{to} vertex. In the example below, loops are removed because loops exist on every vertex in this example, as blood between two people of matching ABO and RhD type can always be exchanged.
<<blood_ex, echo=TRUE, fig.cap= "Directed network of blood type donations in humans.", fig.width=6, fig.height=6>>=
ggplot(data = bloodnet, aes(from_id = from, to_id = to)) +
  geom_net(colour="red", layout="circle", label=TRUE, size=3,
           directed=TRUE, vjust=-0.5, arrowsize=1.5) + theme_net()
@
This network is shown in figure \ref{fig.cap:blood_ex}.  Here, we have used the aesthetics \texttt{vcolour} and \texttt{vsize} set to identity values to change the size and color of all vertices. We have also used the \texttt{layout} and \texttt{vlabel} arguments to change the default layout to a circle layout and to print the blood types, respectively. The circle layout places blood types of the same ABO type next to each other and spreads the vertices out far enough to distinguish between the various ``in" and ``out" types.  You can tell clearly from this plot that the O- type is the universal donor: it has out degree of seven and in degree of zero. Additionally, you can see that the AB+ type is the universal recepient, with in degree of seven and out degree of zero. Anyone looking at this plot can quickly determine which type(s) of blood they can receive and which type(s) can receive their blood.
\begin{figure}
\centering
\includegraphics[width=0.6\textwidth]{figure/blood_ex-1.pdf}
\caption{\label{fig.cap:blood_ex} Network of blood donation possibilities in humans by ABO and RhD blood types.}
\end{figure}
\afterpage{\clearpage}

\subsection{Email Network}
  This email network comes from the 2014 VAST Challenge \citep{emailnet}. It is a directed network of emails between company employees with 55 vertices and 9,063 edges. Each vertex is an employee of the company, and each edge is an email sent from one employee to one or more other employees. The arrow of the directed edge points to the recipient(s) of the email. The network contains two business weeks of emails across the entire company. In order to better visualize the structure of the communication network between employees, all emails that were sent out to all employees are removed in the subsequent examples.
%
\begin{figure}
\centering
<<email_ex, fig.cap = 'All emails within a company over a two week period.',echo=TRUE, out.width='\\textwidth'>>=
ggplot(data = emailnet, aes(from_id = From, to_id = to)) +
  geom_net(aes(colour= CurrentEmploymentType),
           linewidth=0.5, alpha=0.1, size=4) +
  scale_colour_brewer(palette="Set1") + theme_net() +
  theme(legend.position="bottom")
@
\includegraphics[width=\textwidth]{figure/email_ex-1.pdf}
\caption{\label{fig.cap:email_ex} Email network within a company over a two week period.}
\end{figure}
\afterpage{\clearpage}

@@

@@

%\includegraphics[width=8in]{figure/email_ex_facet-1.pdf}
<<email_ex_facet, fig.height=5, fig.width=8, fig.cap = 'All emails within a company over a two week period facetted by day of the month.',echo=TRUE, out.width='\\textwidth'>>=
ggplot(data = emailnet, aes(from_id = From, to_id = to)) +
  geom_net(aes(colour= CurrentEmploymentType), fiteach=TRUE,
           linewidth=0.5, alpha=0.2, size=2) +
  scale_colour_brewer(palette="Set1") + theme_net() +
  theme(legend.position="bottom") + facet_wrap(~day, nrow = 2)
\includegraphics[width=\textwidth]{figure/email_ex_facet-1.pdf}

@@

%' This plot is shown in figure \ref{fig.cap: email_ex_facet}. With the facetting, we can see that there are several days where one or more departments do not communicate with any of the other departments. There are only two days, 13 and 15, without any isolated department communications. Facetting is one of the major benefits of creating a geometry for networks in \texttt{ggplot2}.  Facetting quickly separates dense networks into separate subnetworks for easy visual comparison and  analyses.
%'   \subsection{\texttt{ggplot2} Theme Elements}
%' This example comes from the \texttt{theme()} help page in the \texttt{ggplot2} documentation \citep{ggplot}.  It is a directed network which shows the structure of the inheritance of theme options in the construction of a \texttt{ggplot2} plot.  There are 53 vertices and 36 edges in this network. Each vertex represents one possible theme option. There is an arrow from one theme option to another if the element represented by the \emph{to} vertex inherits its values from the \emph{from} vertex.  For example, the \texttt{axis.ticks.x} option inherits its value from the \texttt{axis.ticks} value, which inherits its value from the \texttt{line} option.  Thus, setting the \texttt{line} option to a value such as \texttt{element\_blank()} sets the entire inheritance tree to \texttt{element\_blank()}, and no lines appear anywhere on the plot background. Finally, we note that the vertices with no edges were incorporated into the plot by adding their labels to the edges data frame in both the  `from\_id' and `to\_id' columns before passing the edges data frame to \texttt{ggplot}.
%' <<theme_ex, echo=TRUE, fig.cap= 'Inheritance structure for \\texttt{theme} elements in \\texttt{ggplot2}.'>>=
%' ggplot(data = theme_elements$edges, aes(from_id = parent, to_id = child)) +
%'   geom_net(vertices = theme_elements$vertices, directed = TRUE, vlabel = TRUE) +
%'   expand_limits(x = c(0,1.2), y = c(0,1)) +
%'   theme_net
%' @
%' \begin{figure}
%' \centering
%' \includegraphics[width=\textwidth]{figure/theme_ex-1.pdf}
%' \caption{\label{fig.cap:theme_ex} Inheritance structure of \texttt{ggplot2} theme elements. This is a recreation of the graph found at \protect\url{http://docs.ggplot2.org/current/theme.html}.}
%' \end{figure}
%' \afterpage{\clearpage}
%'
%' The inheritance structure is plotted in figure \ref{fig.cap:theme_ex}.  In this plot, it is easy to quickly determine the parent and child vertices. Using this plot made creation of the \texttt{theme\_net} object used throughout these examples very simple. We just made set each of the major parent elements, \texttt{text}, \texttt{rect}, and \texttt{line} to \texttt{element\_blank()} and then set the aspect ratio equal to one.
%'   \subsection{Mad Men Networks}
%'   The following code creates the network example given in the introduction. We changed the vertex size and edge color for all vertices and edges, included vertex labels, and colored the vertices according to the character's gender.
%' <<print_ex, ref.label="madmen_ex", eval=FALSE, echo=TRUE,cache=FALSE>>=
%' @
%'  There is another Mad Men network included in the \texttt{gcookbook} package \citep{madmen}. It is a directed network, also of romantic relationships between characters, but it also includes advances made by one character that were rejected by the other.  For example, Roger Sterling made advances toward Betty Draper, but Betty refused him, and so there is a directed edge going from Roger to Betty, but not from Betty to Roger.  If the advance is reciprocated, like between Sal Romano and the Bellhop, there are two directed edges between the two vertices.
%' <<madmen2_ex,echo=TRUE, fig.cap= 'A directed network of relationships in Mad Men.'>>=
%' ggplot(data = mm.directed$edges, aes(from_id = Name1, to_id = Name2)) +
%'   geom_net(vertices = mm.directed$vertices, directed = T, vlabel = T,
%'            vsize = I(2.5), layout = 'fruchtermanreingold',
%'            aes(vcolour=c( "#FF69B4", "#0099ff")[as.numeric(Gender)])) +
%'   expand_limits(x = c(0,1.1), y = c(0,1)) + theme_net
%' @
%' \begin{figure}
%' \centering
%' \includegraphics[width=\textwidth]{figure/madmen2_ex-1.pdf}
%' \caption{\label{fig.cap:madmen2_ex} A directed network of relationships in Mad Men.}
%' \end{figure}
%' This network is shown in figure \ref{fig.cap:madmen2_ex}.  This network is much more connected than the previous Mad Men example.  It also allows us to see much more of the drama in the show. For instance, you can see that Roger made advances towards Betty, his business partner's wife, which is much more scandalous than what we can see in the first network.
%'
%'   \subsection{College Football}
%'   This next example comes from M.E.J. Newman's network data web page \citep{football}.  It is an undirected network consisting of all regular season college football games played between Division I schools in Fall of 2000.  There are 115 vertices and 613 edges: each vertex represents a school, and an edge represents a game played between two schools. There is an additional variable in the vertex data frame corresponding to the conference each team belongs to, and there is an additional variable in the edge data frame that is equal to one if the game occured between teams in the same conference or zero if the game occured between teams in different conferences.
%' <<football_ex, echo=TRUE, fig.cap="All regular season Division I college football games in the Fall 2000 season.">>=
%' colors <- brewer.pal(12, name = 'Paired')
%' ggplot(data = football$edges, aes(from_id = from, to_id = to,
%'                                   elinetype = in.conf + 1)) +
%'   geom_net(vertices = football$vertices, ealpha = 0.3, vsize = 2, vlabel = TRUE,
%'            aes(vcolour=colors[as.numeric(factor(value))])) +
%'   expand_limits(x = c(0,1), y = c(0,1)) + theme_net
%' @
%' The network of football games is given in figure \ref{fig.cap:football_ex}. Here, we have changed the \texttt{elinetype} aesthetic to correspond to games that occur between teams in the same conference or different conferences.  These lines are dotted and solid, respectively. We have also assigned a different color to each conference, and the vertices and their labels are colored according to their conference. This coloring and changing of the linetypes make the structure of the game network easier to view.  There is one conference consisting of Navy, Notre Dame, Utah State, Central Florida, and Connecticut, which is spread out all over the network, whereas most other conferences' teams are all very close to each other because they play within conference much more than they play out of conference.  At the time, these five schools were all independents and did not have a home conference.  Without the coloring capability, we would not have been able to pick out that difference as easily.
%' \begin{figure}
%' \centering
%' \includegraphics[width=\textwidth]{figure/football_ex-1.pdf}
%' \caption{\label{fig.cap:football_ex} The network of regular season Division I college football games in the season of fall 2000. The vertices and their labels are colored by conference.}
%' \end{figure}
%' \afterpage{\clearpage}
%'
%'   \subsection{Les Mis\'{e}rables}
%'   This next network comes from \citet{lesmis}. It is an undirected network of coappearances of characters in Victor Hugo's \emph{Les Mis\'{e}rables}.  There are 77 vertices representing each of the 77 characters in the book, and an edge connects two vertices if those two characters appear in the same chapter of the book.  There are 254 edges in this network. The edges are also weighted by the number of coappearances. The largest weighting is 31, between the characters Jean Valjean and Cosette.  This network is shown in figure \ref{fig.cap:lesmis_ex}. This network reveals many important character groups, as well as the main characters in the story.  The main characters are the ones with high degree, like Valjean, Cosette, and Javert.  There is an extremely wide edge connecting Valjean and Cosette, which implies that they are strongly connected; in fact, Cosette is Valjean's adopted daughter.
%' <<lesmis_ex, echo=TRUE, fig.cap="Coappearance network of characters in Victor Hugo's \\emph{Les Mis\\'{e}rables}">>=
%' ggplot(data = lesmis$edges, aes(from_id = from, to_id = to,
%'                                 esize = value/mean(value))) +
%'   geom_net(vertices = lesmis$vertices, ealpha = 7/10,
%'            vcolour = I("red"), vsize = I(2.5),
%'            layout = 'fruchtermanreingold', vlabel=TRUE) +
